Examining equity in accessibility to bike share: a balanced floating catchment area approach

### 2021-08-30 Bike-share-spatial-equity/Bike-share-spatial-equity-R1.tex
Proof reading Elise's edits before resubmitting R2; created document with marked changes between versions

% Pandoc citation processing

@@

The introduction of PBSPs has been accompanied by a flurry of research
focusing on benefits from them. Many studies have examined differences
in demographics or socioeconomic status between those who use or have
access to PBSPs and those who don't, while other research has explored
spatial inequities in where stations are located (e.g., Hosford and
Winters, 2018; Hull Grasso et al., 2020; Mooney et al., 2019; Qian and
Jaller, 2021, 2020; Smith et al., 2015). In Chicago, Qian and Jaller
(2020) estimated ridership in the city's PBSP and found that a minority
of bike share stations were located in disadvantaged communities, while
annual members from such areas had a lower share of trips compared to
other areas in the city. Similar results were found in Philadelphia.
Despite efforts to increase equity within the city's PBSP, census block
groups with lower median income generated fewer trips (Caspi and Noland,
2019). Trips from stations in such areas were utilitarian (e.g.,
commuting to work), which points to the importance of ensuring equitable
access (Caspi and Noland, 2019). In the case of Seattle, all
neighborhoods had some level of access to dockless bicycles but those
with higher incomes and more residents of higher education had more
bikes (Mooney et al., 2019). Babagoli et al.~(2019) also found that
neighborhoods in New York City with higher affluence had the greatest
proportion of Citi Bike stations.

Chen et al.~(2019) distinguish two types of equity related to bike
share, horizontal and vertical, based on the work of Delbosc and Currie
(Delbosc and Currie, 2011). Horizontal equity leads to balanced or equal
distribution of accessibility and costs for all similar groups, while
vertical equity would involve greater and targeted access for
under-represented or disadvantaged populations (see Chen et al., 2019).
Both are of interest to researchers and transportation planners since
they are often linked in that advantage, or conversely disadvantage, has
spatial patterns. Bike share equity programs, as identified by McNeil et
al.~(2019), can be considered efforts to improve vertical equity since
they favour groups that have benefited less from PBSPs.

On the whole, existing studies highlight the need for PBSPs to be more
accessible for diverse populations in order to increase use beyond the
``typical'' users. This has been the focus of recent research (see,
among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
MacArthur et al., 2020). Offering more people the option of using
sustainable and active transportation, particularly those who have lower
socioeconomic status and might benefit the most, is a worthy policy goal
for cities with PBSPs. However, exploring transportation equity by
investigating where bike share stations are located, often using
neighborhoods or census tracts as the geographical unit of analysis, can
ignore or miss the benefits that may be derived from adjacent zones.
Meaning that, stations may be lacking in certain neighborhoods but there
may be stations accessible within a reasonable walking time. This is
where geographical accessibility becomes an important consideration.

@@

The location and size of docking stations are two factors that are
relevant to accessibility. Since the time or distance needed to reach a
docking station decreases the potential of accessing the program, the
location matters. Kabra et al.~(2020) found that the majority of bike
share usage in Paris, France comes from areas within 300 m of stations,
which amounts to 2-4 minutes walking by an adult who does not have a
disability. Similar to other public amenities affected by crowding, the
utility of docking stations is also limited by the \emph{maximum} number
of bicycles that they can hold (e.g., their size). Accessibility
analyses for PBSPs constitute a positive and evaluation-based approach
that also has the potential to inform equity efforts. For instance, Wang
and Lindsey (2019b) investigated whether new or relocated bike share
stations increased accessibility and use, which offered important
insights to improve the performance of the program.

@@

When measuring accessibility, researchers have also taken different
approaches when it comes to the aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (J. Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals. Chen et al.~(2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation''.

@@

Our analysis builds upon a previous study (Hosford and Winters, 2018),
which found that areas in Hamilton with less advantage are better served
by the city's PBSP compared to other Canadian cities (i.e., Toronto,
Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less
deprived have greater access. Hosford and Winters (2018, p. 47)
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas''. Therefore, the
core service area for the PBSP in Hamilton by default covers more of the
disadvantaged areas in the city. However, there is also a great deal of
variation in income in the core service area because of the local
university and increasing gentrification. Hosford and Winters (2018)
took a macro-level approach in their analysis by using dissemination
areas \emph{across} the city as the unit of analysis. They did not focus
specifically on the core service area and did not differentiate between
conventional and equity stations.

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/hamilton-and-sobi-service-area-1}

@@

{\centering \includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/sobi-stations-in-hamilton-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/interpolated-population-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-7-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-8-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-9-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-10-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-11-1}

@@

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income at the minimum threshold of three minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-3}
\end{figure}

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income at the average threshold of five minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-5}
\end{figure}

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income at the maximum threshold of ten minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-10}
\end{figure}

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income at the extreme threshold of fifteen with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-15}
\end{figure}

@@

Using disaggregated population data, we examined equity in accessibility
to Hamilton Bike Share, with a particular focus on assessing the
contribution of the program's equity stations. The BFCA approach,
combined with pycnophylactic interpolation, enabled us to measure
accessibility on a micro scale which better reflects the scale at which
walking takes place and avoids the ``absorption of disparities'', as
articulated by Chen et al.~(2019) and elsewhere in the transportation
literature (National Academies of Sciences, 2004; Rowangould et al.,
2016). Our method also considers demand and supply in the calculation of
accessibility, which is important for services like PBSPs but not
accounted for in other common accessibility models. Unlike the 2SFCA
method, there was no inflation or deflation in our accessibility
estimates because population and level of service were preserved (see
Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}). This
differentiates our analysis from similar papers exploring equity in
PBSPs that use larger geographical units of analysis or that focus on
station location instead of level of service. In this way, our paper has
made an important contribution by applying an intuitive and useful
approach to measure accessibility to a PBSP. This open and reproducible
research could be expanded for additional analysis using other
individual-level data in Hamilton (e.g., age, gender, household size,
education, etc.) or applied to PBSPs in other North American cities.

The sensitivity analysis revealed that accessibility to docking stations
is maximized at five minutes and decreases significantly by eight
minutes {[}see Table \ref{tab:accessibility-income}{]}. This reflects
the normative guide advertised on some docking stations in Hamilton
showing other stations within a five minute walk, as well as the
directive of NACTO (City Transportation Officials, 2015). We find that
over 118,000 people can access a bike share station within a 15 minute
walk, which represents roughly 85\% of the total population in the core
service area {[}see Table \ref{tab:accessibility-income}{]}. At a
minimum threshold of three minutes, too few people can reach stations
which leads to relatively high levels of service since there is little
crowding for the stations. However, accessibility is at its lowest after
eight minutes whereby congestion effects due to increased potential
demand kick in. Chen et al.~(2019) also found that accessibility
decreases as larger geographical units are used in the estimation, since
more people can access the service.

@@

This paper highlights the importance of both the location and size of
docking stations for increasing equity in accessibility to PBSPs.
Several studies have examined who can access PBSPs based on the location
of stations (among others, see Babagoli et al., 2019; Qian and Jaller,
2020; Qian et al., 2020; J. Wang and Lindsey, 2019b), but the size of
docking stations in disadvantaged communities is not considered in the
literature. Station size (i.e., potential maximum supply of bicycles)
was also not identified by PBSP owners and operators as a factor that
needed to be addressed in equity efforts (Howland et al., 2017). PBSPs
may not necessarily reduce inequities in access if stations are easy to
reach for disadvantaged groups but offer only a small number of
bicycles. Likewise, more people may not opt to use the program if the
supply of bicycles available at nearby stations is insufficient to meet
demand. Babgoli et al.~(2019) found a slight but not statistically
significant increase in the proportion of neighborhoods with the highest
levels of poverty that had stations after the Citi Bike expansion in
2015. Although the Citi Bike expansion was not specifically driven by a
desire to reduce inequity in access, 16\% of neighborhoods with the
highest levels of poverty had stations compared to 12\% before. This
leaves open the question as to whether the stations held enough bicycles
to meet demand, and whether congestion effects in neighbourhoods with
more poverty are reducing the potential of bike share trips.

The following policy and planning implications can inform efforts for
increasing equity in Hamilton Bike Share, and potentially other PBSPs in
North America. Our analysis has identified specific areas that have both
low accessibility and low median total household income. Figures
\ref{fig:figure-bi-map-threshold-3},
\ref{fig:figure-bi-map-threshold-5},
\ref{fig:figure-bi-map-threshold-10}, and
\ref{fig:figure-bi-map-threshold-15} highlight potential locations for
new equity stations to better accommodate low income groups. Additional
stations are needed to further reduce inequities in accessibility. This
information will be useful to transportation planners in Hamilton and
HBS operators. Our findings also suggest that PBSPs should consider
increasing the size of stations in disadvantaged communities. Docking
stations added to target vertical equity should have an adequate supply
of public bicycles, particularly more in areas that have a strong
interest in adopting bike share. Finally, distance to stations is
associated with bike share use (Fuller et al., 2011; J. Wang and
Lindsey, 2019b), especially when the trip starts at the residence
(Fishman et al., 2013). Distance can also be a barrier to using PBSPs
(Fishman et al., 2014). To address this issue, transportation planners
and PBSP operators should consult with groups and communities who are
under-represented among bike share users to determine how far they are
willing to walk to reach a docking station and to get their input on the
location of new docking stations.

@@

Wang and Lindsey (2019b) have noted that there is a lack of research
that examines how bike share users' behaviour changes as a result of
program changes to station locations or improvements in accessibility.
As such, a logical next step to this research is to examine whether the
equity stations increased ridership or resulted in new memberships in
areas that were previously under-serviced. An examination of the types
of trips undertaken by residents in these areas would also be
informative, such as the study undertaken by Caspi and Norland (2019)
after bike share stations were implemented in low-income Philadelphia
neighborhoods. The bulk of cycling facilities that have been built in
Hamilton to date are located in the core service area near the
conventional stations. It would be worthwhile to explore the route
choice of bike share trips departing or ending at the equity stations
and to identify factors that specifically influence trips from these
stations, which would extend existing studies conducted by Scott and
colleagues (Lu et al., 2018; Scott and Ciuro, 2019; Scott et al., 2021).
This paper, combined with additional studies such as those
conceptualized above, would serve as a valuable case study for Hamilton
and other cities with PBSPs that wish to evaluate and address inequities
in accessibility and transportation options in urban areas.

@@

We wish to thank the anonymous reviewers who provided constructive
feedback to improve the quality of our paper. This research was
completed using open software, and we also wish to acknowledge the
developers of the following \texttt{R} packages: \texttt{biscale}
(Prener et al., 2020), \texttt{cowplot} (Wilke, 2020),
\texttt{data.table} (Dowle and Srinivasan, 2021), \texttt{disk.frame}
(ZJ, 2021), \texttt{gdistance} (van Etten, 2020), \texttt{gridExtra}
(Auguie, 2017), \texttt{kableExtra} (Zhu, 2021), \texttt{knitr} (Xie,
2021a), \texttt{pycno} (Brunsdon, 2014), \texttt{r5r} (Saraiva et al.,
2021), \texttt{raster} (Hijmans, 2020), \texttt{rgdal} (Bivand et al.,
2021), \texttt{rticles} (Allaire et al., 2021), \texttt{sf} (Pebesma,
2021), \texttt{tidyverse} (Wickham, 2021), \texttt{tinytex} (Xie,
2021b), \texttt{units} (Pebesma et al., 2021).

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-R-rticles}{}%
Allaire, J., Xie, Y., R Foundation, Wickham, H., Journal of Statistical
Software, Vaidyanathan, R., Association for Computing Machinery,
Boettiger, C., Elsevier, Broman, K., Mueller, K., Quast, B., Pruim, R.,
Marwick, B., Wickham, C., Keyes, O., Yu, M., Emaasit, D., Onkelinx, T.,
Gasparini, A., Desautels, M.-A., Leutnant, D., MDPI, Taylor and Francis,
Ögreden, O., Hance, D., Nüst, D., Uvesten, P., Campitelli, E.,
Muschelli, J., Hayes, A., Kamvar, Z.N., Ross, N., Cannoodt, R., Luguern,
D., Kaplan, D.M., Kreutzer, S., Wang, S., Hesselberth, J., Dervieux, C.,
2021. Rticles: Article formats for r markdown.

\leavevmode\hypertarget{ref-auchinclossDesignBaselineDescription2020}{}%
Auchincloss, A.H., Michael, Y.L., Fuller, D., Li, S., Niamatullah, S.,
Fillmore, C.E., Setubal, C., Bettigole, C., 2020. Design and baseline
description of a cohort of bikeshare users in the city of Philadelphia.
Journal of Transport \& Health 16, 100836.
doi:\href{https://doi.org/10.1016/j.jth.2020.100836}{10.1016/j.jth.2020.100836}

\leavevmode\hypertarget{ref-R-gridExtra}{}%
Auguie, B., 2017. GridExtra: Miscellaneous functions for "grid"
graphics.

\leavevmode\hypertarget{ref-babagoliExploringHealthSpatial2019}{}%
Babagoli, M.A., Kaufman, T.K., Noyes, P., Sheffield, P.E., 2019.
Exploring the health and spatial equity implications of the New York
City Bike share system. Journal of Transport \& Health 13, 200--209.
doi:\href{https://doi.org/10.1016/j.jth.2019.04.003}{10.1016/j.jth.2019.04.003}

@@

\leavevmode\hypertarget{ref-breyWantRideMy2017}{}%
Brey, R., Castillo-Manzano, J.I., Castro-Nuño, M., 2017. ``I want to
ride my bicycle'': Delimiting cyclist typologies. Applied Economics
Letters 24, 549--552.
doi:\href{https://doi.org/10.1080/13504851.2016.1210760}{10.1080/13504851.2016.1210760}

@@

\leavevmode\hypertarget{ref-buckAreBikeshareUsers2013}{}%
Buck, D., Buehler, R., Happ, P., Rawls, B., Chung, P., Borecki, N.,
2013. Are Bikeshare Users Different from Regular Cyclists?: A First Look
at Short-Term Users, Annual Members, and Area Cyclists in the
Washington, D.C., Region. Transportation Research Record 2387, 112--119.
doi:\href{https://doi.org/10.3141/2387-13}{10.3141/2387-13}

>>>>

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

The introduction of PBSPs has been accompanied by a flurry of research
focusing on benefits from them. Many studies have examined differences
in demographics or socioeconomic status between those who use or have
access to PBSPs and those who don't, while other research has explored
spatial inequities in where stations are located (e.g., Hosford and
Winters, 2018; Mooney et al., 2019; Qian and Jaller, 2021, 2020; Smith
et al., 2015; \textbf{hullgrassoBikeShareEquity2020?}). In Chicago, Qian
and Jaller (2020) estimated ridership in the city's PBSP and found that
a minority of bike share stations were located in disadvantaged
communities, while annual members from such areas had a lower share of
trips compared to other areas in the city. Similar results were found in
Philadelphia. Despite efforts to increase equity within the city's PBSP,
census block groups with lower median income generated fewer trips
(Caspi and Noland, 2019). Trips from stations in such areas were
utilitarian (e.g., commuting to work), which points to the importance of
ensuring equitable access (Caspi and Noland, 2019). In the case of
Seattle, all neighborhoods had some level of access to dockless bicycles
but those with higher incomes and more residents of higher education had
more bikes (Mooney et al., 2019). Babagoli et al. (2019) also found that
neighborhoods in New York City with higher affluence had the greatest
proportion of Citi Bike stations.

Chen et al. (2019) distinguish two types of equity related to bike
share, horizontal and vertical, based on the work of Delbosc and Currie
(Delbosc and Currie, 2011). Horizontal equity leads to balanced or equal
distribution of accessibility and costs for all similar groups, while
vertical equity would involve greater and targeted access for
under-represented or disadvantaged populations (see Chen et al., 2019).
Both are of interest to researchers and transportation planners since
they are often linked in that advantage, or conversely disadvantage, has
spatial patterns. Bike share equity programs, as identified by McNeil et
al. (2019), can be considered efforts to improve vertical equity since
they favour groups that have benefited less from PBSPs.

On the whole, existing studies highlight the need for PBSPs to be more
accessible for diverse populations in order to increase use beyond the
``typical'' users. This has been the focus of recent research (see,
among others, Auchincloss et al., 2020; MacArthur et al., 2020;
\textbf{hullgrassoBikeShareEquity2020?}). Offering more people the
option of using sustainable and active transportation, particularly
those who have lower socioeconomic status and might benefit the most, is
a worthy policy goal for cities with PBSPs. However, exploring
transportation equity by investigating where bike share stations are
located, often using neighborhoods or census tracts as the geographical
unit of analysis, can ignore or miss the benefits that may be derived
from adjacent zones. Meaning that, stations may be lacking in certain
neighborhoods but there may be stations accessible within a reasonable
walking time. This is where geographical accessibility becomes an
important consideration.

@@

The location and size of docking stations are two factors that are
relevant to accessibility. Since the time or distance needed to reach a
docking station decreases the potential of accessing the program, the
location matters. Kabra et al. (2020) found that the majority of bike
share usage in Paris, France comes from areas within 300 m of stations,
which amounts to 2-4 minutes walking by an adult who does not have a
disability. Similar to other public amenities affected by crowding, the
utility of docking stations is also limited by the \emph{maximum} number
of bicycles that they can hold (e.g., their size). Accessibility
analyses for PBSPs constitute a positive and evaluation-based approach
that also has the potential to inform equity efforts. For instance, Wang
and Lindsey (2019b) investigated whether new or relocated bike share
stations increased accessibility and use, which offered important
insights to improve the performance of the program.

@@

When measuring accessibility, researchers have also taken different
approaches when it comes to the aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals. Chen et al. (2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation.''

@@

Our analysis builds upon a previous study (Hosford and Winters, 2018),
which found that areas in Hamilton with less advantage are better served
by the city's PBSP compared to other Canadian cities (i.e., Toronto,
Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less
deprived have greater access. Hosford and Winters (2018, p. 47)
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas.'' Therefore, the
core service area for the PBSP in Hamilton by default covers more of the
disadvantaged areas in the city. However, there is also a great deal of
variation in income in the core service area because of the local
university and increasing gentrification. Hosford and Winters (2018)
took a macro-level approach in their analysis by using dissemination
areas \emph{across} the city as the unit of analysis. They did not focus
specifically on the core service area and did not differentiate between
conventional and equity stations.

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/hamilton-and-sobi-service-area-1}

@@

{\centering \includegraphics[width=1\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/sobi-stations-in-hamilton-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/interpolated-population-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-7-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-8-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-9-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-10-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-11-1}

@@

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income at the minimum threshold of three minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-3}
\end{figure}

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income at the average threshold of five minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-5}
\end{figure}

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income at the maximum threshold of ten minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-10}
\end{figure}

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income at the extreme threshold of fifteen with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-15}
\end{figure}

@@

Using disaggregated population data, we examined equity in accessibility
to Hamilton Bike Share, with a particular focus on assessing the
contribution of the program's equity stations. The BFCA approach,
combined with pycnophylactic interpolation, enabled us to measure
accessibility on a micro scale which better reflects the scale at which
walking takes place and avoids the ``absorption of disparities,'' as
articulated by Chen et al. (2019) and elsewhere in the transportation
literature (National Academies of Sciences, 2004; Rowangould et al.,
2016). Our method also considers demand and supply in the calculation of
accessibility, which is important for services like PBSPs but not
accounted for in other common accessibility models. Unlike the 2SFCA
method, there was no inflation or deflation in our accessibility
estimates because population and level of service were preserved (see
Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}). This
differentiates our analysis from similar papers exploring equity in
PBSPs that use larger geographical units of analysis or that focus on
station location instead of level of service. In this way, our paper has
made an important contribution by applying an intuitive and useful
approach to measure accessibility to a PBSP. This open and reproducible
research could be expanded for additional analysis using other
individual-level data in Hamilton (e.g., age, gender, household size,
education, etc.) or applied to PBSPs in other North American cities.

The sensitivity analysis revealed that accessibility to docking stations
is maximized at five minutes and decreases significantly by eight
minutes {[}see Table \ref{tab:accessibility-income}{]}. This reflects
the normative guide advertised on some docking stations in Hamilton
showing other stations within a five minute walk, as well as the
directive of NACTO (City Transportation Officials, 2015). We find that
over 118,000 people can access a bike share station within a 15 minute
walk, which represents roughly 85\% of the total population in the core
service area {[}see Table \ref{tab:accessibility-income}{]}. At a
minimum threshold of three minutes, too few people can reach stations
which leads to relatively high levels of service since there is little
crowding for the stations. However, accessibility is at its lowest after
eight minutes whereby congestion effects due to increased potential
demand kick in. Chen et al. (2019) also found that accessibility
decreases as larger geographical units are used in the estimation, since
more people can access the service.

@@

This paper highlights the importance of both the location and size of
docking stations for increasing equity in accessibility to PBSPs.
Several studies have examined who can access PBSPs based on the location
of stations (among others, see Babagoli et al., 2019; Qian and Jaller,
2020; Wang and Lindsey, 2019b;
\textbf{qianEnhancingEquitableService2020?}), but the size of docking
stations in disadvantaged communities is not considered in the
literature. Station size (i.e., potential maximum supply of bicycles)
was also not identified by PBSP owners and operators as a factor that
needed to be addressed in equity efforts (Howland et al., 2017). PBSPs
may not necessarily reduce inequities in access if stations are easy to
reach for disadvantaged groups but offer only a small number of
bicycles. Likewise, more people may not opt to use the program if the
supply of bicycles available at nearby stations is insufficient to meet
demand. Babgoli et al. (2019) found a slight but not statistically
significant increase in the proportion of neighborhoods with the highest
levels of poverty that had stations after the Citi Bike expansion in
2015. Although the Citi Bike expansion was not specifically driven by a
desire to reduce inequity in access, 16\% of neighborhoods with the
highest levels of poverty had stations compared to 12\% before. This
leaves open the question as to whether the stations held enough bicycles
to meet demand, and whether congestion effects in neighbourhoods with
more poverty are reducing the potential of bike share trips.

The following policy and planning implications can inform efforts for
increasing equity in Hamilton Bike Share, and potentially other PBSPs in
North America. Our analysis has identified specific areas that have both
low accessibility and low median total household income. Figures
\ref{fig:figure-bi-map-threshold-3},
\ref{fig:figure-bi-map-threshold-5},
\ref{fig:figure-bi-map-threshold-10}, and
\ref{fig:figure-bi-map-threshold-15} highlight potential locations for
new equity stations to better accommodate low income groups. Additional
stations are needed to further reduce inequities in accessibility. This
information will be useful to transportation planners in Hamilton and
HBS operators. Our findings also suggest that PBSPs should consider
increasing the size of stations in disadvantaged communities. Docking
stations added to target vertical equity should have an adequate supply
of public bicycles, particularly more in areas that have a strong
interest in adopting bike share. Finally, distance to stations is
associated with bike share use (Fuller et al., 2011; Wang and Lindsey,
2019b), especially when the trip starts at the residence (Fishman et
al., 2013). Distance can also be a barrier to using PBSPs (Fishman et
al., 2014). To address this issue, transportation planners and PBSP
operators should consult with groups and communities who are
under-represented among bike share users to determine how far they are
willing to walk to reach a docking station and to get their input on the
location of new docking stations.

@@

Wang and Lindsey (2019b) have noted that there is a lack of research
that examines how bike share users' behaviour changes as a result of
program changes to station locations or improvements in accessibility.
As such, a logical next step to this research is to examine whether the
equity stations increased ridership or resulted in new memberships in
areas that were previously under-serviced. An examination of the types
of trips undertaken by residents in these areas would also be
informative, such as the study undertaken by Caspi and Norland (2019)
after bike share stations were implemented in low-income Philadelphia
neighborhoods. The bulk of cycling facilities that have been built in
Hamilton to date are located in the core service area near the
conventional stations. It would be worthwhile to explore the route
choice of bike share trips departing or ending at the equity stations
and to identify factors that specifically influence trips from these
stations, which would extend existing studies conducted by Scott and
colleagues (Lu et al., 2018; Scott et al., 2021; Scott and Ciuro, 2019).
This paper, combined with additional studies such as those
conceptualized above, would serve as a valuable case study for Hamilton
and other cities with PBSPs that wish to evaluate and address inequities
in accessibility and transportation options in urban areas.

@@

We wish to thank the anonymous reviewers who provided constructive
feedback to improve the quality of our paper. This research was
completed using open software, and we also wish to acknowledge the
developers of the following \texttt{R} packages: \texttt{biscale}
(Prener et al., 2020), \texttt{cowplot} (Wilke, 2020),
\texttt{data.table} (Dowle and Srinivasan, 2021), \texttt{disk.frame}
(ZJ, 2021), \texttt{gdistance} (van Etten, 2020), \texttt{gridExtra}
(Auguie, 2017), \texttt{kableExtra} (Zhu, 2021), \texttt{knitr} (Xie,
2021a), \texttt{pycno} (Brunsdon, 2014), \texttt{r5r} (Saraiva et al.,
2021), \texttt{raster} (Hijmans, 2021), \texttt{rgdal} (Bivand et al.,
2021), \texttt{rticles} (Allaire et al., 2021), \texttt{sf} (Pebesma,
2021), \texttt{tidyverse} (Wickham, 2021), \texttt{tinytex} (Xie,
2021b), \texttt{units} (Pebesma et al., 2021).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-R-rticles}{}%
Allaire, J., Xie, Y., R Foundation, Wickham, H., Journal of Statistical
Software, Vaidyanathan, R., Association for Computing Machinery,
Boettiger, C., Elsevier, Broman, K., Mueller, K., Quast, B., Pruim, R.,
Marwick, B., Wickham, C., Keyes, O., Yu, M., Emaasit, D., Onkelinx, T.,
Gasparini, A., Desautels, M.-A., Leutnant, D., MDPI, Taylor and Francis,
Ögreden, O., Hance, D., Nüst, D., Uvesten, P., Campitelli, E.,
Muschelli, J., Hayes, A., Kamvar, Z.N., Ross, N., Cannoodt, R., Luguern,
D., Kaplan, D.M., Kreutzer, S., Wang, S., Hesselberth, J., Dervieux, C.,
2021. Rticles: Article formats for r markdown.

\leavevmode\hypertarget{ref-auchinclossDesignBaselineDescription2020}{}%
Auchincloss, A.H., Michael, Y.L., Fuller, D., Li, S., Niamatullah, S.,
Fillmore, C.E., Setubal, C., Bettigole, C., 2020. Design and baseline
description of a cohort of bikeshare users in the city of
{Philadelphia}. Journal of Transport \& Health 16, 100836.
doi:\href{https://doi.org/10.1016/j.jth.2020.100836}{10.1016/j.jth.2020.100836}

\leavevmode\hypertarget{ref-R-gridExtra}{}%
Auguie, B., 2017. gridExtra: Miscellaneous functions for "grid"
graphics.

\leavevmode\hypertarget{ref-babagoliExploringHealthSpatial2019}{}%
Babagoli, M.A., Kaufman, T.K., Noyes, P., Sheffield, P.E., 2019.
Exploring the health and spatial equity implications of the {New York
City Bike} share system. Journal of Transport \& Health 13, 200--209.
doi:\href{https://doi.org/10.1016/j.jth.2019.04.003}{10.1016/j.jth.2019.04.003}

@@

\leavevmode\hypertarget{ref-breyWantRideMy2017}{}%
Brey, R., Castillo-Manzano, J.I., Castro-Nuño, M., 2017. 'I want to ride
my bicycle': Delimiting cyclist typologies. Applied Economics Letters
24, 549--552.
doi:\href{https://doi.org/10.1080/13504851.2016.1210760}{10.1080/13504851.2016.1210760}

@@

\leavevmode\hypertarget{ref-buckAreBikeshareUsers2013}{}%
Buck, D., Buehler, R., Happ, P., Rawls, B., Chung, P., Borecki, N.,
2013. Are bikeshare users different from regular cyclists?: A first look
at short-term users, annual members, and area cyclists in the
washington, d.c., region. Transportation Research Record 2387, 112--119.
doi:\href{https://doi.org/10.3141/2387-13}{10.3141/2387-13}
